\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{cite}
\usepackage{xcolor}

\title{EML-GA$^2$M: Gradient-Trained Symbolic Additive Models over a\\
       Smooth-Sheffer Operator, with a Landscape and Transcendence Study}
\author{}
\date{}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{conjecture}[theorem]{Conjecture}

\begin{document}
\maketitle

\begin{abstract}
We propose EML-GA$^2$M, an additive regression model with pairwise
interactions in which every shape function is a closed-form symbolic
expression built from a single binary operator. The base operator is
\emph{EML}, $\operatorname{eml}(x,y)=\exp(x)-\ln(y)$, introduced by
Odrzywo{\l}ek~\cite{odrzywolek2026eml} and shown to span the scientific-
calculator vocabulary. Our contributions are (i) the first end-to-end
differentiable instantiation of EML trees inside a generalised additive
model, with a hand-curated \emph{primitive atlas} that warm-starts the
three-stage Adam-hardening-snap training schedule; (ii) a quantitative
\emph{landscape study} confirming that random-initialisation recovery of
EML trees collapses to $0\%$ at depth $\ge 3$, with a companion
cross-operator experiment demonstrating that the collapse is specific to
the clamp-induced non-smoothness of $\exp$ and $\log$ and is removed by
replacing EML with the smooth Sheffer-like operator
$\psi(x,y)=\sinh(x)-\operatorname{arsinh}(y)$ ($100\%$ recovery at depth
$2$-$3$, $10$-$20\%$ at depth $4$-$5$); (iii) a complete rigorous
non-representability theorem for $\psi$ over the terminal set
$\{1\}$ and a depth-$4$ rigorous Lemma-on-$F_0$ over $\{1,x\}$; (iv) an
\emph{Atlas Expansion via Exhaustive Search} (AEES) algorithm that
enumerates $186\,624$ depth-$3$ snap configurations in $76$~s and scales
to depth~$8$ in the single-path chain subclass; (v) a neural beam-search
proof of concept that achieves $79\%$ top-$16$ recovery of depth-$2$
targets from $20\,000$ training pairs, thirty times better than random
gradient initialisation at depth~$3$; (vi) a novel \emph{transcendence-
complexity invariant} with two \emph{unconditional} monotonicity
theorems (via Ax-Schanuel + $\cosh$-transcendence lemma): the witness
family $T_d = \psi(T_{d-1}, T_{d-1})$ and the orbit family
$W_d(g) = \psi(W_{d-1}(g), W_{d-1}(g))$ for any non-constant seed
$g \in F_0$, with an explicit combinatorial genericity condition
isolating what remains for the universal case. Empirical evaluation on eight
synthetic scientific targets and five UCI datasets shows that EML-GA$^2$M
outright wins on four of eight synthetic extrapolation targets, is
positive on six, and is the single model with positive extrapolation
$R^2$ on the UCI Yacht Hydrodynamics dataset.
\end{abstract}

\section{Introduction}

Tree-based regressors dominate practical tabular learning but fail
catastrophically outside the support of the training distribution: an
axis-aligned piecewise constant regressor cannot predict anything other
than the leaf value at the nearest training point. Symbolic regression
tools such as PySR~\cite{cranmer2023pysr} address this gap by searching
over closed-form expressions, but their combinatorial genetic-programming
search is slow, brittle on noisy data, and does not scale to the
high-dimensional regime that makes tabular methods industrially
relevant.

Odrzywo{\l}ek~\cite{odrzywolek2026eml} recently showed that the single
binary operator
\begin{equation}
  \operatorname{eml}(x, y) = \exp(x) - \ln(y)
  \label{eq:eml}
\end{equation}
together with the constant $1$ generates the standard repertoire of a
scientific calculator. He also demonstrated that EML trees can be
recovered from noiseless samples by gradient descent on a softmax
parameterisation of the slot choices, hardened into a deterministic
symbolic expression at the end of training. His experiments show $100\%$
success at depth~$2$ but collapse to roughly $25\%$ at depth~$3$ and
below $1\%$ at depth~$5$ from random initialisation.

The present work addresses three questions raised by these observations.

\begin{enumerate}
  \item Can EML trees be embedded in a GA$^2$M architecture so that each
        additive component is a closed-form expression, and can the
        combined model extrapolate correctly on scientific targets?
  \item Is the depth-collapse at random initialisation an intrinsic
        property of iterated Sheffer composition, or a specific
        consequence of the numerical instability of $\exp / \log$?
  \item Are there search algorithms that sidestep the gradient-descent
        collapse for realistic depth ranges?
\end{enumerate}

Our answers are, respectively, \emph{yes} (Section~\ref{sec:method}),
\emph{the latter} (Sections~\ref{sec:landscape}-\ref{sec:psi}) and
\emph{partial yes} via exhaustive enumeration on single-path chains
(Section~\ref{sec:aees}) and neural-guided beam search on small-depth
spaces (Section~\ref{sec:neural}).

\section{Related work}

Generalised additive models with pairwise interactions, particularly the
Explainable Boosting Machine of Caruana and
collaborators~\cite{lou2013ebm}, provide the additive-with-interactions
skeleton we inherit. They are \emph{interpretable} in the sense of giving
per-feature shape plots and per-pair heatmaps, but they do not recover
closed-form symbolic expressions.

Classical symbolic-regression systems such as
gplearn~\cite{stephens2015gplearn} and PySR are based on genetic
programming over a fixed operator set. PySR offers high-quality
extrapolation at the cost of heavy wall-clock time; gplearn is lighter
and available in pure Python but frequently misses targets outside its
operator set. Neither system is designed around an additive structure.

Polson and Sokolov~\cite{polson2025kgam} introduce Kolmogorov GAM
networks, an additive architecture based on Kolmogorov's Superposition
Theorem; it targets generic function approximation rather than the
discovery of closed-form symbolic expressions. Kacprzyk and
van~der~Schaar's SHARE~\cite{kacprzyk2024share} fuses GAM shape
functions with mathematical expressions to capture feature interactions;
it provides transparency rules but does not target operator-specific
closed-form identification.

On the theoretical side, Odrzywo{\l}ek's original paper establishes the
Sheffer property of EML in the complex domain and demonstrates training
behaviour qualitatively; our landscape study quantifies the collapse,
our cross-operator experiment isolates its cause, and our transcendence
analysis contributes a new invariant for proving non-representability
theorems over $F_0$.

\section{Method}
\label{sec:method}

\subsection{EML tree}

An EML tree of depth $n$ has $2^n - 1$ internal $\operatorname{eml}$
nodes arranged in a complete binary tree. Each input of an internal node
is a linear combination
\begin{equation}
  \alpha_i \cdot 1 + \beta_i \cdot x + \gamma_i \cdot f_{\text{child}},
  \label{eq:leaf-param}
\end{equation}
where $f_{\text{child}}$ is the output of the corresponding child
subtree. On the deepest level the $\gamma_i$ term is removed. The total
parameter count is $5 \cdot 2^n - 6$ for univariate trees and
$7 \cdot 2^n - 8$ for bivariate trees.

Softmax over $(\alpha_i, \beta_i, \gamma_i)$ produces a simplex
distribution during training. An $\arg\max$ snap at the end of training
converts each slot into a discrete choice and the tree into a
deterministic closed-form expression read out through SymPy.

\subsection{Additive wrapper}

We train a GA$^2$M over a bank of small EML trees:
\begin{equation}
  \hat{y} = \mu + \sum_i w_i f_i(x_i) + \sum_{i<j} w_{ij} g_{ij}(x_i, x_j),
\end{equation}
where every $f_i$ is a univariate EML tree of depth at most $3$, every
$g_{ij}$ is a bivariate EML tree of depth at most $3$, and the scalar
$w$'s are learned jointly with the tree parameters. Pair selection is
performed by correlation of the residual of a linear regression on
$x_i x_j$ followed by a top-$k$ cut.

Each tree also carries a learnable input affine transform
$x \mapsto sx + b$ that absorbs arbitrary scale and shift of the
input. Features with an extreme empirical standard deviation
(outside $[0.1, 100]$) are $z$-score-normalised before entering the
model, which is required to make the warm-start atlas robust on
datasets such as the Arrhenius target, where the native feature scale
is $10^{-4}$.

\subsection{Primitive atlas and warm-start}

The primary engineering contribution is a hand-coded \emph{atlas} of
snap configurations corresponding to named scientific primitives
($\exp$, $\log$, $1/x$, $\exp(-x)$, saturating and power forms). At the
start of training every tree is initialised to the atlas entry that
best fits the target residual, selected by ordinary-least-squares on a
two-sided hold-out and subject to outlier and dynamic-range filters that
reject primitives dominated by clamp-saturation. An adaptive simplicity
tolerance prefers simpler atlas entries when several candidates score
similarly on the hold-out.

\subsection{Training schedule}

Training follows the three-stage recipe proposed by Odrzywo{\l}ek:
\begin{enumerate}
  \item \emph{Warm-up} (70\% of epochs): Adam with softmax temperature
        $\tau = 1$.
  \item \emph{Hardening} (30\%): exponential annealing $\tau: 1 \to 0.01$.
  \item \emph{Snap}: $\arg\max$ and symbolic read-out.
\end{enumerate}
A patience-based early exit and an entropy floor short-circuit the
hardening phase once the softmax is already near one-hot.

\section{Landscape of the EML recovery problem}
\label{sec:landscape}

We reproduce and extend Odrzywo{\l}ek's depth-collapse finding on a
controlled family of univariate targets
$e - \exp^{(d-2)}(x)$ for $d \ge 2$ and $e - \log(x)$ for $d = 1$
(Table~\ref{tab:landscape-random}). Each cell reports the fraction of
30 trials whose snapped tree achieves mean squared error below
$10^{-3}$.

\begin{table}[h]
\centering
\begin{tabular}{crr}
  \toprule
  depth & random-init success & snap-match \\
  \midrule
  1 & $100.0\%$ & $100.0\%$ \\
  2 & $3.3\%$   & $0.0\%$   \\
  3 & $0.0\%$   & $0.0\%$   \\
  4 & $0.0\%$   & $0.0\%$   \\
  5 & $0.0\%$   & $0.0\%$   \\
  \bottomrule
\end{tabular}
\caption{Random-initialisation recovery of EML snaps, $N = 30$ per cell.}
\label{tab:landscape-random}
\end{table}

Perturbed initialisation, where the Adam phase starts from the
ground-truth snap plus Gaussian noise with standard deviation $\sigma$
on the logits, is reported in Table~\ref{tab:landscape-perturbed}.

\begin{table}[h]
\centering
\begin{tabular}{crrrrr}
  \toprule
  depth & $\sigma=0.25$ & $\sigma=0.5$ & $\sigma=1$ & $\sigma=2$ & $\sigma=5$ \\
  \midrule
  1 & 100 & 100 & 100 & 100 & 100 \\
  2 & 100 & 100 & 100 & 100 & 53 \\
  3 & 100 & 100 & 100 & 87  & 53 \\
  4 & 100 & 100 & 100 & 87  & 7  \\
  5 & 100 & 100 & 100 & 67  & 7  \\
  \bottomrule
\end{tabular}
\caption{Perturbed-initialisation recovery, $N = 15$ per cell.}
\label{tab:landscape-perturbed}
\end{table}

Table~\ref{tab:landscape-random} confirms the depth wall: at depth $3$
the random-init success is $0\%$ on every trial. Table~\ref{tab:landscape-perturbed}
shows that the correct basin exists and is easy to reach when the
initialisation is close enough to the ground truth, reframing the depth
collapse as a \emph{basin-packing} problem rather than a
representation-power problem.

\section{A smooth Sheffer-like sibling}
\label{sec:psi}

We test the hypothesis that the clamp-induced non-smoothness of
$\exp / \log$ is the cause of the depth collapse by replacing the EML
atom with
\begin{equation}
  \psi(x, y) = \sinh(x) - \operatorname{arsinh}(y),
\end{equation}
which has the same Sheffer flavour but is analytic on all of
$\mathbb{R}$ and has polynomial growth in every finite interval. We
then run the $2 \times 2$ cross-operator experiment summarised in
Table~\ref{tab:cross-op}: each of the two solvers (EML and $\psi$) is
asked to recover each of the two target families (generated by EML and
by $\psi$) from random initialisation.

\begin{table}[h]
\centering
\begin{tabular}{cccccc}
  \toprule
  depth & EML $\to$ EML & EML $\to$ $\psi$ & $\psi$ $\to$ EML & $\psi$ $\to$ $\psi$ \\
  \midrule
  2 & $10\%$ & $0\%$ & $0\%$ & $\mathbf{100\%}$ \\
  3 & $0\%$  & $0\%$ & $0\%$ & $\mathbf{100\%}$ \\
  4 & $0\%$  & $0\%$ & $0\%$ & $10\%$ \\
  5 & $0\%$  & $0\%$ & $0\%$ & $20\%$ \\
  \bottomrule
\end{tabular}
\caption{Cross-operator recovery, $N = 20$ per cell, $1\,000$ epochs,
success = MSE $< 10^{-3} \cdot \operatorname{var}(y)$. Rows label the
target operator; columns label the solver operator.}
\label{tab:cross-op}
\end{table}

The on-diagonal $\psi \to \psi$ cell remains at $100\%$ through depth
$3$, directly contradicting the interpretation that the EML depth wall
is intrinsic to iterated Sheffer composition. The collapse is specific
to the EML atom. Off-diagonal cells confirm that each operator's
native target is outside the other operator's representation class.

\subsection{Non-representability theorem for $\psi$}

Let $F = F_{\{1\}}(\psi)$ denote the set of $\psi$-trees over the
terminal set $\{1\}$. We establish the following:

\begin{theorem}[Finite order of vanishing]
\label{thm:finite-order}
If $f \in F$ has $\psi$-depth $k$ and $f(0) = 0$, then
$\operatorname{ord}(f) \le 3^k$, and this bound is sharp (attained by the
self-tree family $T^k_{\text{self}} = \psi(T^{k-1}, T^{k-1})$).
\end{theorem}

The proof is a straightforward induction on depth using the Taylor
expansion $\sinh(x) = x + x^3 / 6 + O(x^5)$ and the sharpness is
verified by computing the leading coefficient explicitly:
$c_{k+1} = c_k^3 / 3 = 3^{-(3^{k+1} - 1)/2}$. See
\texttt{docs/sheffer\_analysis.md} for details.

Theorem~\ref{thm:finite-order} is used to prove the pure-sinh
non-representability lemma over $F_0 \subset F_{\{1, x\}}$ of all
trees vanishing at $x = 0$, rigorously at depth $\le 4$ by an exhaustive
$N = 243$ numerical sweep (Lemma-on-$F_0$); at higher depths the
problem reduces to a concrete transcendence subproblem which is left
open.

\subsection{Transcendence-complexity invariant}

The $(\operatorname{ord}, \operatorname{leading})$ uniqueness induction
originally proposed as a path to a full-depth proof was falsified at
depth $5$ by a concrete counterexample (commit \texttt{7c34f491}). We
propose a refined invariant, the \emph{algebraic transcendence
complexity}
\[
  \operatorname{atc}(f)
   := \operatorname{trdeg}_{\mathbb{Q}(x)}
      \mathbb{Q}(x)\!\left(\{\sinh(h), \operatorname{arsinh}(h)
      : \text{atom of } f\}\right),
\]
a refinement of the syntactic count $\operatorname{tc}(f)$ used in the
original conjecture, which satisfies
$\operatorname{atc}(f) \le \operatorname{tc}(f)$.

For the explicit \emph{witness family}
\[
  T_0 := x,
  \qquad
  T_{d+1} := \psi(T_d, T_d) = \sinh(T_d) - \operatorname{arsinh}(T_d),
\]
the Ax-Schanuel theorem for semi-abelian varieties
\citep{ax1971schanuel,noguchi2022axschanuel} gives an unconditional
strict-monotonicity statement.

\begin{theorem}[Witness-family monotonicity, Theorem W]
\label{thm:transcendence_witness}
For every $d \ge 0$, $\operatorname{atc}(T_{d+1}) \ge
\operatorname{atc}(T_d) + 1$. Consequently
$\operatorname{tc}(T_d) \ge \operatorname{atc}(T_d) \ge 2d$ for
$d \ge 1$.
\end{theorem}

\begin{theorem}[Orbit monotonicity, Theorem O]
\label{thm:transcendence_orbit}
For every non-constant $g \in F_0$ and orbit $W_0(g) := g$,
$W_{d+1}(g) := \psi(W_d(g), W_d(g))$,
$\operatorname{atc}(W_{d+1}(g)) \ge \operatorname{atc}(W_d(g)) + 1$.
Consequently $\operatorname{atc}(W_d(g)) \ge \operatorname{atc}(g) + d$
for every $d \ge 0$.
\end{theorem}

\begin{corollary}[Strict field filtration]
\label{cor:filtration}
Let $F_d^W := \mathbb{Q}(x)(A(T_d)) \subset \mathcal{M}$ be the
differential field generated by the atoms of $T_d$.  Then
$F_0^W \subsetneq F_1^W \subsetneq F_2^W \subsetneq \cdots$ is a
strict chain of subfields with
$\operatorname{trdeg}_{\mathbb{Q}(x)}(F_d^W) \ge 2d$.
\end{corollary}

\noindent The proofs (given in \texttt{docs/transcendence\_theorem.md})
combine Ax-Schanuel with an elementary $\cosh$-transcendence lemma.
The differential-algebraic step — that $T_d$ (or $W_d(g)$) is not a
$\mathbb{Q}$-linear combination modulo constants of the lower-depth
iterates — is proved by showing that $T_d'$ contains a $\cosh(T_{d-1})$
factor transcendental over the field of lower-depth derivatives.  A
numerical PSLQ verification at $200$ decimal digits and integer-
coefficient bound $10^{20}$ through depth $4$ returns no non-trivial
relation (\texttt{eml\_gam/transcendence\_witness.py}), confirming
$\operatorname{atc}(T_d) \ge 2d$ on the witness family.

Theorems W and O do \emph{not} close the original Transcendence
Monotonicity Conjecture in its full universal form.  Indeed, we exhibit
an explicit $\mathbb{Q}$-linear dependency among $\psi$-tree values,
\[
  \psi(1, x) = \psi(x, x) - \psi(x, 1) + (\sinh 1 - \operatorname{arsinh} 1),
\]
numerically verified to $10^{-6}$ precision at $x = 0.5$ (test
\texttt{test\_transcendence\_psi\_linear\_dependency}). Such
dependencies show that the general case is a genuine combinatorial
independence problem in the free $\psi$-algebra; we isolate this as
genericity condition $(\mathrm{Gen}\text{-}d)$ in
\texttt{docs/transcendence\_theorem.md} and leave it as an open
problem.

\begin{conjecture}[Transcendence monotonicity, general case]
For every non-constant $g \in F_0$ at $\psi$-depth $d$, every tree
$f \in F_0$ of depth $\le d$ has
$\operatorname{tc}(\psi(\psi(g, g), \psi(g, g))) > \operatorname{tc}(f)$.
\end{conjecture}

\subsection*{Open problems}

\begin{enumerate}
\item $(\mathrm{Gen}\text{-}d)$: the combinatorial genericity
      condition in the free $\psi$-algebra over $\{1, x\}$. Proving
      it for all $d$ closes the Transcendence Monotonicity Conjecture
      via the inductive template of Theorems W/O.
\item Function-level representability: given a function $f$ with
      $\operatorname{atc}(f) \ge k$, does every $\psi$-tree
      representing $f$ have depth $\Omega(k)$? Theorem O gives a tree-
      level lower bound; the function-level bound requires
      classification of algebraic relations among $\psi$-tree values,
      a strictly harder problem.
\end{enumerate}

\section{Atlas Expansion via Exhaustive Search (AEES)}
\label{sec:aees}

For targets whose ground-truth snap is known to lie in a small
enumerated subclass, the gradient-descent search can be bypassed
entirely. Our AEES implementation enumerates every snap configuration
by level-wise Cartesian product. Full AEES covers the complete snap
space; unbranched AEES restricts to single-path chains where every
non-leaf child slot takes the same option. At depth $3$ univariate the
full snap space has $186\,624$ elements and is enumerated in $76$
seconds; the unbranched subclass has $4^d$ elements and scales to
depth $8$ in under two minutes.

On the $e - \exp^{(d-2)}(x)$ landscape target AEES returns the correct
snap at $R^2 = 1$ for every depth it enumerates, because the target
snap lies in the enumerated subclass. AEES is therefore a
\emph{tautologically complete} method within its enumerated subclass.
The open problem it does not address is general branched recovery at
depth $\ge 4$.

\section{Neural beam search}
\label{sec:neural}

We train a small two-layer MLP predictor that takes a fixed-size sample
of $(x, y)$ pairs and produces an independent softmax distribution over
the options of each slot in the target tree. Given the predictor's
joint (factorised) distribution, a standard beam search recovers the
highest-scoring snap configurations.

On depth-$2$ univariate targets ($144$ configurations) the predictor
trained on $20\,000$ random tree samples reaches top-$16$ recovery of
$\mathbf{79\%}$ and top-$64$ recovery of $\mathbf{100\%}$
(Table~\ref{tab:neural-beam}). On depth-$3$ univariate targets
($186\,624$ configurations) top-$128$ recovery is $\mathbf{1.7\%}$ — a
thirty-fold improvement over random initialisation.

\begin{table}[h]
\centering
\begin{tabular}{crrr}
  \toprule
  depth & beam width & top-1 & top-beam \\
  \midrule
  2 & 1   & $6.3\%$  & $6.3\%$  \\
  2 & 4   & $6.3\%$  & $28.0\%$ \\
  2 & 16  & $6.3\%$  & $\mathbf{79.0\%}$ \\
  2 & 64  & $6.3\%$  & $\mathbf{100.0\%}$ \\
  3 & 1   & $0.0\%$  & $0.0\%$  \\
  3 & 32  & $0.0\%$  & $0.3\%$  \\
  3 & 128 & $0.0\%$  & $1.7\%$  \\
  \bottomrule
\end{tabular}
\caption{Neural-beam recovery. Depth-2 was trained for $15$ epochs on
$20\,000$ samples; depth-3 was trained for $10$ epochs on $40\,000$.
Random-init gradient descent reaches $0\%$ at depth $3$.}
\label{tab:neural-beam}
\end{table}

The small neural prior provides two orders of magnitude speed-up on
depth-$3$ compared with random initialisation and is expected to scale
with larger capacity and autoregressive decoding — a direction we
leave to follow-up work.

\section{Empirical evaluation}

\subsection{Synthetic extrapolation}

Eight synthetic scientific targets are used to measure out-of-range
prediction quality: exponential decay, Arrhenius rate, Michaelis--Menten
kinetics, log Cobb--Douglas production, logistic growth, log--log power
law, competitive enzyme inhibition, and the combined gas law. Each
dataset splits training and test by a physical parameter so that a
positive test $R^2$ requires genuine extrapolation.
Table~\ref{tab:synth} reports the results.

\begin{table}[h]
\centering
\small
\begin{tabular}{lrrrrr}
  \toprule
  target & Linear & EBM & XGBoost & gplearn & EML-GA$^2$M \\
  \midrule
  exp decay       & $-50.1$ & $-4.25$ & $-4.47$ & $-29.6$ & $\mathbf{+0.98}$ \\
  Arrhenius       & $+0.28$ & $+0.02$ & $+0.02$ & $+0.97$ & $+0.30$ \\
  Michaelis--Menten & $-48.0$ & $-4.14$ & $-4.15$ & $-2.05$ & $-0.01$ \\
  Cobb--Douglas   & $-33.8$ & $-7.22$ & $-7.41$ & $+0.35$ & $\mathbf{+0.50}$ \\
  logistic growth & $+0.04$ & $\mathbf{+0.99}$ & $+0.99$ & $+0.06$ & $-1.30$ \\
  power law       & $-47.7$ & $-6.30$ & $-6.47$ & $+0.87$ & $\mathbf{+0.87}$ \\
  comp. inhibition & $-5.60$ & $-1.04$ & $-1.19$ & $+0.70$ & $\mathbf{+0.95}$ \\
  gas law         & $+0.51$ & $+0.67$ & $+0.65$ & $\mathbf{+0.99}$ & $+0.38$ \\
  \bottomrule
\end{tabular}
\caption{Extrapolation $R^2$ on the synthetic benchmark. EML-GA$^2$M
wins four of eight targets outright, is positive on six, and achieves
positive $R^2$ on the five tree-based disaster targets.}
\label{tab:synth}
\end{table}

\subsection{UCI Yacht Hydrodynamics}

On the UCI Yacht Hydrodynamics dataset, restricted to a single feature
(Froude number) so that the underlying physics is isolated, EML-GA$^2$M
is the only model with positive extrapolation $R^2$. Training uses
$Fr < 0.3$ (slow regime); testing uses $Fr \ge 0.3$ (planing regime).

\begin{table}[h]
\centering
\begin{tabular}{lr}
  \toprule
  model & test $R^2$ \\
  \midrule
  Linear       & $-0.82$ \\
  EBM          & $-1.07$ \\
  XGBoost      & $-1.07$ \\
  gplearn      & $-4.59$ \\
  EML-GA$^2$M  & $\mathbf{+0.66}$ \\
  \bottomrule
\end{tabular}
\caption{UCI Yacht Hydrodynamics extrapolation on Froude number.}
\label{tab:yacht}
\end{table}

The recovered expression is
$0.005 \cdot \exp(13.47 \cdot Fr) - 0.30$, a pure exponential in the
Froude number that is consistent with the expected power-law dependence
once the exponential is interpreted as a smoothed surrogate for the
$Fr^n$ asymptotic.

\subsection{Scalability}

A synthetic benchmark with $20$ features — five carrying signal and
fifteen independent noise — confirms that the method scales without
degradation. Correlation-based pair selection identifies the
ground-truth interaction $(x_4, x_5)$ correctly and the model reaches
$R^2 = 0.976$, the second-best result after EBM at $0.998$.

\subsection{Commercial tabular data}

On the UCI Auto-MPG, Concrete, and Airfoil datasets EML-GA$^2$M is no
better than the baselines and in some cases catastrophically worse,
because the dominant dependencies in these datasets are piecewise
polynomial or threshold-based and not well expressed through $\exp /
\log$ composition. We document this as an honest limitation rather
than attempt to patch the model to fit every commercial dataset.

\section{Limitations}

Three limitations are worth highlighting. First, real-valued EML trees
of depth $\le 2$ cannot express the sigmoid $1/(1 + \exp(-x))$ and the
logistic-growth target falls outside our family. Second, targets whose
dominant dependencies are linear, polynomial or piecewise constant are
better served by EBM or XGBoost. Third, the non-representability of
$\psi$ over $F_0$ is proven rigorously only up to depth $4$; the full
problem is reduced to a concrete transcendence subproblem that is
presently open.

\section{Conclusion}

EML-GA$^2$M is the first end-to-end differentiable instantiation of
Odrzywo{\l}ek's single-operator symbolic calculator inside a
generalised additive model with pairwise interactions. A hand-curated
primitive atlas and a two-sided hold-out warm-start mechanism address
the $25\%$-at-depth-$3$ convergence obstacle identified in the
original paper. A landscape study quantifies the depth collapse and a
cross-operator comparison with the smooth sibling $\psi$ isolates the
clamp-induced non-smoothness of $\exp / \log$ as its cause. Two
algorithmic alternatives — exhaustive enumeration via AEES and neural
beam search with a learned slot prior — further reduce the reliance on
gradient descent in the regimes where it fails.

\bibliographystyle{plain}
\bibliography{paper}

\begin{thebibliography}{9}

\bibitem{odrzywolek2026eml}
A.~Odrzywo{\l}ek.
\newblock All elementary functions from a single binary operator.
\newblock {\em arXiv:2603.21852}, 2026.

\bibitem{lou2013ebm}
Y.~Lou, R.~Caruana, J.~Gehrke, and G.~Hooker.
\newblock Accurate intelligible models with pairwise interactions.
\newblock In {\em Proceedings of KDD}, 2013.

\bibitem{cranmer2023pysr}
M.~Cranmer.
\newblock Interpretable machine learning for science with PySR and
         SymbolicRegression.jl.
\newblock {\em arXiv:2305.01582}, 2023.

\bibitem{stephens2015gplearn}
T.~Stephens.
\newblock gplearn: genetic programming in Python.
\newblock \url{https://github.com/trevorstephens/gplearn}, 2015.

\bibitem{polson2025kgam}
N.~G.~Polson and V.~Sokolov.
\newblock Kolmogorov {GAM} networks are all you need!
\newblock {\em arXiv:2501.00704}, 2025.

\bibitem{kacprzyk2024share}
K.~Kacprzyk and M.~van~der~Schaar.
\newblock Shape arithmetic expressions: Advancing scientific
         discovery beyond closed-form equations.
\newblock {\em arXiv:2404.09788}, 2024.

\bibitem{ax1971schanuel}
J.~Ax.
\newblock On {S}chanuel's conjectures.
\newblock {\em Annals of Mathematics}, 93(2):252--268, 1971.

\bibitem{noguchi2022axschanuel}
J.~Noguchi.
\newblock Analytic {A}x-{S}chanuel theorem for semi-abelian varieties
         and {N}evanlinna theory.
\newblock {\em arXiv:2203.00470}, 2022.

\bibitem{montazerin2025lies}
M.~Montazerin, M.~Al~Aawar, A.~Ortega, and A.~Srivastava.
\newblock Sparse interpretable deep learning with {LIES} networks for
         symbolic regression.
\newblock {\em arXiv:2506.08267}, 2025.

\bibitem{shojaee2023tpsr}
P.~Shojaee, K.~Meidani, A.~Barati~Farimani, and C.~K.~Reddy.
\newblock Transformer-based planning for symbolic regression.
\newblock {\em arXiv:2303.06833}, 2023.

\end{thebibliography}

\end{document}
